---
title: "Simulation - RSV study evaluating RSVpreF vs IG"
subtitle: "Sequential analyses"
author: "maj"
date: "2025-03-28"
date-modified: last-modified
bibliography: refs.bib
csl: biomed-central.csl
embed-resources: true
---

```{r}  
#| label: setup
#| code-summary: Libraries and globals
#| code-fold: true

suppressWarnings(suppressPackageStartupMessages(library(cmdstanr)))
library(ggplot2)
library(parallel)
library(mcprogress)
suppressWarnings(suppressPackageStartupMessages(library(tictoc)))

library(poisson)
library(data.table)
library(survival)
library(flexsurv)
library(extraDistr)

tic()

set.seed(92818785)

mc_cores <- 60

tok <- unlist(tstrsplit(getwd(), split = "/"))
if(tok[length(tok)] == "rsv-grant"){
  mod_01 <- cmdstanr::cmdstan_model("./stan/log-logistic-aft-01.stan")
} else {
  mod_01 <- cmdstanr::cmdstan_model("../stan/log-logistic-aft-01.stan")
}

output_dir_mcmc <- paste0(getwd(), "/tmp")
output_dir_res <- paste0(getwd(), "/out")

```

## Background

### Aim

We want to compare effectiveness of IG vs RSVpreF for protection against RSV in the first 360 days of life.

### Interventions

+ IG - monoclonal antibody given to infant at birth
+ RSVpreF - maternal vaccine given after 32 weeks gestation

### Randomisation

Balanced 1:1 randomisation. 
Variations might be warranted under the use of historical controls.

### Primary outcome

Options:

+ binary outcome variable indicating medically attended per the definition of MATISSE study:

  + MA-RTI visit AND RSV-positive test result AND one or more of the following
    + Fast breathing (RR $\ge 60$  bpm for $<2$  months of age [$<60$ days of age], $\ge 50$ bpm for $2 - <12$ months of age, or $\ge 40$ bpm for 12–24 months of age)
    + SpO2 $<95%$
    + Chest wall indrawing

### Estimand

Primary: population perspective, i.e. the estimated comparative effectiveness of each implemented strategy with adherence as observed under real-world conditions.

### Follow up

Follow up to age 360 days for both groups.

### Prior information

Weakly regularising priors are adopted as soft-constraints on all model parameters.

### Resource Constraints

Sample size up to approximately 3000 dyads.

## Methods

Here we consider a single interim analysis assessing futility and superiority as stopping criteria followed by a final analysis at the maximum sample size.
Variations on this adaptation might consider sample size re-estimation.

## Operating characteristics

### Reference design

A reference design with single analysis at 3000 dyads has been considered previously.

### Sequential design


```{r}


l_res <- qs::qread(file.path(output_dir_res, "20250401184925-sim-res-keep.qs"))


d_sim_dat <- rbindlist(lapply(1:length(l_res), function(i){
  l_res[[i]]$d
}), idcol = "id_sim")

d_sim_res <- rbindlist(lapply(1:length(l_res), function(i){
  l_res[[i]]$d_pars
}), idcol = "id_sim")

d_sim_res[, sup := fifelse(pr_drmst_gt_0 > p_sup, 1, 0)]
d_sim_res[, fut := fifelse(pr_drmst_gt_0 < p_fut, 1, 0)]
# 
# d_sim_res[id_analy == 2, N_enrol := 2500]
# d_sim_res[id_analy == 3, N_enrol := 3000]

# The static imputation allows us to get unbiased summaries over all interim
# analyses. Because of the potential for early stopping, later interim analyses
# do not reflect a random selection and therefore the summary results would be
# biased.
d_sim_res[, sup := nafill(sup, type = "locf"), keyby = id_sim]
d_sim_res[, fut := nafill(fut, type = "locf"), keyby = id_sim]
d_sim_res[, mu_rmst0 := nafill(mu_rmst0, type = "locf"), keyby = id_sim]
d_sim_res[, mu_rmst1 := nafill(mu_rmst1, type = "locf"), keyby = id_sim]
d_sim_res[, mu_drmst := nafill(mu_drmst, type = "locf"), keyby = id_sim]
d_sim_res[, p0_360 := nafill(p0_360, type = "locf"), keyby = id_sim]
d_sim_res[, p1_360 := nafill(p1_360, type = "locf"), keyby = id_sim]
d_sim_res[, rd_360 := nafill(rd_360, type = "locf"), keyby = id_sim]
d_sim_res[, p0_sim := nafill(p0_sim, type = "locf"), keyby = id_sim]
d_sim_res[, p1_sim := nafill(p1_sim, type = "locf"), keyby = id_sim]
d_sim_res[, rd_sim := p1_sim - p0_sim]

d_sim_res[, .(id_sim, id_analy, p0_sim, p1_sim, 
              rd_sim, rd_360, 
              mu_rmst0, mu_rmst1, pr_drmst_gt_0,
              sup, fut)]


# plot survival curves based on expectation of posterior means
d_fig <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, 
                       p0_sim, p1_sim, p0_360, p1_360,
                       sup, fut)]

d_fig[p0_360 < 0.075 & id_analy == 3 & sup != 1]
d_fig[id_sim == 2]
ggplot(d_fig[id_analy == 3],
       aes(x = 0, y = p0_360)) +
  geom_point()





dd <- d_sim_dat[id_sim == 5]
dd[, t_fu := max(d_c$t_0) + 360 - t_0]
dd[, evt := as.numeric(t_fu >= t_evt & t_evt <= 360)]
# Ensure that the time we use in the analysis stops at the 360 days
dd[, t_evt_obs := copy(t_evt)]
# Censored events are set to 360 days
dd[evt == 0, t_evt_obs := 360]

km1 <- survfit(Surv(t_evt_obs, evt) ~ trt, data = dd)
# Visualize with survminer
ggsurvplot(km1, data = dd, risk.table = F,
           break.y.by = 0.025,
           conf.int = F,   
           ggtheme = theme_minimal(),
           ylim = c(0.85, 1))

# trt has to be a factor otherwise flexsurvreg will only plot overall
dd[, trt := factor(trt)]
fs1 <- flexsurvreg(Surv(t_evt_obs, evt) ~ trt, data = dd, dist = "llogis")
print(fs1)
str(fs1)
ggflexsurvplot(fs1, data = dd, risk.table = F,
               break.y.by = 0.025,
               conf.int = F,   
               ggtheme = theme_minimal(),
               ylim = c(0.85, 1))

# Survival probability at day 360
pllogis(360, shape = exp(fs1$opt$par[1]), scale = exp(fs1$opt$par[2]))
pllogis(360, shape = exp(fs1$opt$par[1]), scale = exp(fs1$opt$par[2] + fs1$opt$par[3]))

#

# Prepare data for Stan
ld <- list(
  N = nrow(dd),
  P = 2,
  X = cbind(1, dd$trt),
  y = dd$t_evt_obs,
  event = dd$evt,
  N_pred = 361,
  t_surv = 0:360,
  # hyperparameters
  mu0_gamma = c(5, 0),
  sd0_gamma = c(5, 2),
  rho_shape = 0.5
)

m1 <- mod_01$sample(
  ld,
  init = function() {list(
    gamma = c(runif(1, 5, 10), runif(1, -1, 1)),
    shape = runif(1, 0, 4)
    )},
  iter_warmup = 1000, iter_sampling = 1000,
  parallel_chains = 1, chains = 1, refresh = 0, show_exceptions = T,
  max_treedepth = 10, adapt_delta = 0.95)

print(fs1)
m1$summary(variables = c("gamma", "shape", "scale0"))
m1$summary(variables = c("p0_360", "p1_360"))

d_post <- data.table(
  m1$draws(
    variables = c("surv0", "surv1"),
    format = "matrix"))

d_fig <- melt(d_post, measure.vars = names(d_post))  
d_fig[variable %like% "surv0", trt := 0]
d_fig[variable %like% "surv1", trt := 1]
d_fig[, day := gsub("surv.\\[", "", variable)]
d_fig[, day := as.numeric(gsub("\\]", "", day)) - 1]
    
d_fig <- d_fig[, .(
  mu_surv = mean(value),
  q_025 = quantile(value, prob = 0.025),
  q_975 = quantile(value, prob = 0.975)
), keyby = .(trt, day)]
d_fig[, trt := factor(trt)]

ggplot(d_fig, aes(x = day, y = mu_surv, group = trt, col = trt))  +
  geom_line(lwd = 0.2) +
  scale_x_continuous(limits = c(0, 400)) +
  scale_y_continuous(
    "Survival probability",
    limits = c(0.85, 1), breaks = seq(0.85, 1, by = 0.025)) +
  theme_minimal() +
  theme(legend.position = "top")
#


d_fig <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, rd_sim, sup, fut)]

d_fig[, mean(sup), keyby = .(N, rd_sim)]
p0 <- sort(unique(d_sim_res$p0_sim))
stopifnot(length(p0) == 1)
p1_uniq <- sort(unique(d_sim_res$p1_sim))
d_fig[, rd_sim := factor(rd_sim, levels = p1_uniq - p0)]

d_fig <- rbind(
  cbind(decision = "Superiority", d_fig[, .(pr_dec = mean(sup)), keyby = .(N, rd_sim)]),
  cbind(decision = "Futility", d_fig[, .(pr_dec = mean(fut)), keyby = .(N, rd_sim)])
)
d_fig[, decision := factor(decision, levels = c("Superiority", "Futility"))]

ggplot(
  d_fig, 
  aes(x = N, y = pr_dec, group = rd_sim, col = rd_sim)) +
  geom_line() +
  geom_point(size = 0.5) +
  scale_color_discrete("RD (day 360)") +
  scale_x_continuous("", c(2000, 2500, 3000)) +
  scale_y_continuous(
    "Cumulative probability of decision", limits = c(0, 1),
    breaks = seq(0, 1, by = 0.1)) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)
  ) +
    guides(col=guide_legend(ncol=1)) +
  facet_wrap(~decision, labeller = label_both) +
  ggtitle("Cumulative probability of decision by enrolment size:")




d_fig <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, rd_sim, rd_360, sup, fut)]
d_fig <- melt(
  d_fig[
    , .(N, rd_sim, rd_360)], 
  measure.vars = c("rd_360"),
  value.name = "mu_rd")
d_fig[, rd_sim := factor(rd_sim, levels = p1_uniq - p0)]

d_fig <- d_fig[, .(
  mu_rd = mean(mu_rd),
  q_025 = quantile(mu_rd, prob = 0.025),
  q_975 = quantile(mu_rd, prob = 0.975)
), keyby = .(N, rd_sim)]

ggplot(
  d_fig, 
  aes(x = N, y = mu_rd)) +
  geom_linerange(aes(ymin = q_025, ymax = q_975), lwd = 0.3) +
  geom_point(size = 0.4) +
  scale_x_continuous("", c(2000, 2500, 3000)) +
  scale_y_continuous(
    "Distribution of mean RD") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)
  ) +
  guides(fill=guide_legend(nrow=4)) +
  facet_wrap(~rd_sim, labeller = label_both)

```





```{r}
#| label: tidy-up
#| code-summary: Shut down timer
#| code-fold: true

toc(log = TRUE, quiet = TRUE)
```

Simulations run: `r unlist(tic.log(format = TRUE))`.

<!-- ::: {#refs} -->
<!-- ::: -->





