---
title: "Simulation - RSV study evaluating RSVpreF vs IG"
author: "maj"
date: "2025-03-19"
date-modified: last-modified
bibliography: refs.bib
csl: biomed-central.csl
embed-resources: true
---

```{r}  
#| label: setup
#| code-summary: Libraries and globals
#| code-fold: true

suppressWarnings(suppressPackageStartupMessages(library(cmdstanr)))
library(ggplot2)
library(parallel)
library(mcprogress)
suppressWarnings(suppressPackageStartupMessages(library(tictoc)))

library(data.table)

tic()

set.seed(92818785)
# Simulation controls
n_sim <- 10000

# Assume somewhat higher risk on the RSVpreF arm than observed in the MATISSE study
# given that the study was run during covid
p_0 <- 0.1

p_1 <- p_0 + seq(-0.03, 0.03, len = n_sim)

# Effects based on risk difference and/or VE
rd <- p_1 - p_0
ve <- 1-(p_1/p_0)


N_per_arm <- 1500

mc_cores <- 60

m1 <- cmdstanr::cmdstan_model("../stan/historic-control-1.stan")
m2 <- cmdstanr::cmdstan_model("../stan/ordinal-01.stan")
m3 <- cmdstanr::cmdstan_model("../stan/ordinal-02.stan")
m4 <- cmdstanr::cmdstan_model("../stan/logistic.stan")
```

## Background

### Aim

We want to compare effectiveness of IG vs RSVpreF for protection against RSV in the first 360 days of life.

### Interventions

+ IG - monoclonal antibody given to infant at birth
+ RSVpreF - maternal vaccine given after 32 weeks gestation

### Randomisation

Balanced 1:1 randomisation. 
Variations might be warranted under the use of historical controls.

### Primary outcome

Options:

+ binary outcome variable indicating medically attended per the definition of MATISSE study:

  + MA-RTI visit AND RSV-positive test result AND one or more of the following
    + Fast breathing (RR $\ge 60$  bpm for $<2$  months of age [$<60$ days of age], $\ge 50$ bpm for $2 - <12$ months of age, or $\ge 40$ bpm for 12–24 months of age)
    + SpO2 $<95%$
    + Chest wall indrawing

+ ordinal variable indicating health state at each follow up:

  + no rsv - no occurrence of positive rsv result (irrespective of medical attendance status)
  + ma - medical attendance for rti with positive rsv
  + hosp - hospitalisation for rti with positive rsv
  + icu - icu for rti with positive rsv
  + death - all cause mortality
  
The ordinal perspective is considered as a method to improve efficiency.
The categories aim to represent a exhaustive and mutually exclusive ranked set.
An additional category for positive rsv but no medical attendance might be added for completeness but it was unclear if this group would have any units so it has been omitted for now.

### Estimand

Primary: population perspective, i.e. the estimated comparative effectiveness of each implemented strategy with adherence as observed under real-world conditions.

### Follow up

Follow up to age 360 days for both groups.

### Prior information

MATISSE study (run during COVID-19) is the primary source of information for day 360 outcomes for RSVpreF (maternal vaccine).
For the medically attended lower RTI with RSV positive test (RSV-MA-LRTI, see Tables S1 and S6 in MATISSE supplementary authored by Kampmann, i.e. interim analysis), there were $92/3495 \approx 2.6%$ cases on RSVpreF arm and $156/3480 \approx 4.5%$ cases on the placebo arm, $VE \approx 0.41$ at day 360.

No day 360 data available for immunoprophylaxis (nirsevimab).

During the COVID-19 pandemic rsv rates were lower than typical and so a baseline risk of 10% is used here.

### Constraints

Sample size up to approximately 3000 dyads.

## Methods

A high-level description of the analysis options include:

+ Conjugate beta binomial analysis comparing the number of binary events by arm at a cross sectional timepoint (e.g. 360 days)
+ Leverage historical controls via partial pooling
+ Ordinal analysis - proportional odds

Other alternatives might include longitudinal analyses on ordinal health states at key follow up times.
Sequential study might enable early stopping.

Here we consider a single analysis at the maximum follow up.

## Operating characteristics

### Reference design

As a reference design, assume the least informative binary indicator with single analysis at maximum sample size (3000 dyads).

Analysis based on conjugate beta binomial model.

Consider two possible trial conclusion criteria, tailored dependening on what is more relevant.
These are arbitrary and applied indepdently, just to motivate discussion.

1. Conclude that IG is more effective if there is a 97.5% (or higher) probability that it reduces the risk of MA-RTI relative to the RSVpreF arm. That is, $Pr(p_1 < p_0) > 0.975 \implies \text{trial success}$. The control arm (RSVpreF) is assumed to have an event risk of $p_0 = 0.1$ and the IG group event risk $p_1 \in \{x : 0.025 \le x \le 0.075\}$ where values lower than $p_0$ imply benefit associated with IG relative to RSVpreF.
2. Conclude that IG is more effective if there is a 90% (or higher) probability that the VE, defined as 1 - RR, is above 20%. That is, $Pr(1 - (p_1 / p_0) > 0.2) > 0.9 \implies \text{trial success}$. VE is computed from the same values nominated as above. When evaluating two active vaccines, it is possible for VE to be less than zero, e.g. when the interventional arm increases risk, $p_1 > p_0$. This approach is simply stating that there needs to be some minimal non-zero effect present, but does not require such a stringent decision criteria.

The power curve is based on interpolating the results from `r n_sim` simulated trials.

```{r}
#| label: oc-1
#| code-summary: Operating characteristics for least efficient design
#| code-fold: true
    
    
y_0 <- rbinom(n_sim, N_per_arm, p_0)
y_1 <- rbinom(n_sim, N_per_arm, p_1)

i <- 1

# Define the function to integrate
integrand_1 <- function(
    x, alpha_0, beta_0,
    alpha_1, beta_1) {
  pbeta(x, alpha_0, beta_0) * dbeta(x, alpha_1, beta_1)
}
  
# Say we are interested in VE  = 1 - (p1/p0) and specifically in the probability
# that Pr(VE > 0.2) = Pr(1 - p1/p0 > 0.2) = Pr(p1/p0 < 0.8).

# We need to integrate over all possible values of p_0 and p_1 considering only the
# cases where p_1/p_0 < 0.8. So we need a double integral
# Pr(p1/p0 < 0.8) = \int_0^1 \int_0^{0.8p_0} f_1(p_1) f_0(p_0) dp_1 dp_0
# This integral accumulates the probability that, given fixed p_0, p_1 falls below 
# 0.8 * p_0
integrand_2 <- function(
    p0, alpha_0, beta_0,
    alpha_1, beta_1, delta = 0) {
  sapply(p0, function(p0_val) {
    integrate(function(p1) dbeta(p1, alpha_1, beta_1), 
              lower = 0, 
              upper = delta * p0_val)$value * dbeta(p0_val, alpha_0, beta_0)
  })
}

res <- pmclapply(1:n_sim, function(i){
  
  # i = sample(1:n_sim, 1)
  # Evaluate whether risk of event is lower on IG arm
  pr_1_lt_0 <- integrate(
    integrand_1, lower = 0, upper = 1,
    alpha_0 = y_1[i] + 1 , 
    beta_0 = N_per_arm - y_1[i] + 1 ,
    alpha_1 = y_0[i] + 1, 
    beta_1 = N_per_arm - y_0[i] + 1)$value
  
  pr_ve_gt_delta <- integrate(
    integrand_2, lower = 0, upper = 1,
    alpha_0 = y_0[i] + 1  , 
    beta_0 = N_per_arm - y_0[i] + 1 ,
    alpha_1 = y_1[i] + 1, 
    beta_1 = N_per_arm - y_1[i] + 1,
    delta = 1 - 0.2)$value
  
  # p_post_0 <- rbeta(1e5, y_0[i] + 1, N_per_arm - y_0[i] + 1)
  # p_post_1 <- rbeta(1e5, y_1[i] + 1, N_per_arm - y_1[i] + 1)
  # c(pr_1_lt_0_int, mean(p_post_1 < p_post_0))
  # ve <- 1 - (p_post_1/p_post_0)
  # c(pr_ve_gt_delta, mean(ve > 0.2))
  
  # # equivalent to testing risk difference against null hypothesis of zero
  # pr_1_lt_0_approx <- mean(p_post_1 < p_post_0)
  
  # assume a typical frequentist critria
  c(pr_1_lt_0 = pr_1_lt_0, 
    pr_ve_gt_delta = pr_ve_gt_delta)
  
  
}, mc.cores = mc_cores)

m_res <- do.call(rbind, res)

d_fig_1 <- data.table(
  x_rd = rd,
  x_ve = ve,
  win_rd = as.numeric(m_res[, 1] > 0.975),
  win_ve = as.numeric(m_res[, 2] > 0.9)
)
```

```{r}
#| echo: FALSE
#| label: fig-pwr_1-rd
#| fig-cap: 'Power curves characterising power by effect size (risk difference) at a sample size of 1500 per arm in two group study with binary outcome'
#| fig.height: 4.5
#| fig.width: 4.5
#| fig-pos: H


ggplot(d_fig_1, aes(x = x_rd, y = win_rd)) +
  geom_smooth(se = F, lwd = 0.3, col = 1, 
              method = "gam", formula = y ~ s(x, bs = "ps")) +
  scale_y_continuous("Power", limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  scale_x_continuous("Risk difference (p1-p0)", breaks = seq(-0.03, 0.03, by = 0.01)) +
  theme_minimal()
```


```{r}
#| echo: FALSE
#| label: fig-pwr_1-ve
#| fig-cap: 'Power curves characterising power by effect size (vaccine effectiveness) at a sample size of 1500 per arm in two group study with binary outcome'
#| fig.height: 4.5
#| fig.width: 4.5
#| fig-pos: H


ggplot(d_fig_1, aes(x = x_ve, y = win_ve)) +
  geom_smooth(se = F, lwd = 0.3, col = 1, 
              method = "gam", formula = y ~ s(x, bs = "ps")) +
  scale_y_continuous("Power", limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  scale_x_continuous("Vaccine effectiveness (1 - RR)", breaks = seq(-1, 0.4, by = 0.2)) +
  theme_minimal()
```



### Historic controls



```{r}
#| label: setup2
#| code-summary: Configuring inputs for simulation
#| code-fold: true
    
n_sim = 3000

# Only look at subset of original p_1
p_1_sub <- quantile(p_1, probs = c(0, 0.25, 0.5, 0.75, 1))

# Prior for treatment group (IG)
# Convert beta parameterisation from mean and sample size to a and b
alpha <- function(mu, nu){
  mu * nu
} 
beta <- function(mu, nu){
  (1-mu)*nu
}

# set mean to 
mu <- 0.05
nu <- 10.05

j = 1

# Effects based on risk difference and/or VE
rd_sub <- p_1_sub - p_0
ve_sub <- 1-(p_1_sub/p_0)

N_ctl <- 1500
N_trt <- 1500
```

Incorporating historic controls can be down within a power-prior framework.
There are lots of ways to implement power-priors, @Ibrahim2015.
Here, the partial pooling approach is used:

$$
\begin{aligned}
Y_0 &\sim \text{Bin}(n_0, p_0) \quad \text{Current control arm RSVpreF} \\
Y_1 &\sim \text{Bin}(n_1, p_1) \quad \text{Current treatment arm IG} \\
Y_h &\sim \text{Bin}(n_h, p_h) \quad \text{Historic data for RSVpreF from trial } h \\
\text{logit}(p_0), \text{logit}(p_h) &\sim \text{N}(\mu, \tau) \quad \text{Common distribution for controls}\\
\mu &\sim \text{Normal}(\text{logit}(0.05), 0.25) \\
\tau &\sim \text{Exp}(3) \\
p_1 &\sim \text{Beta}(0.5, 9.5)
\end{aligned}
$$

$\mu$ and $\tau$ represent the between study mean and SD. 
When $\tau$ is small, all the logits are similar, motivating greater use of the historic information.
When $\tau$ is large, we have observed high variability in the control rate and thus we do not want to incorporate much of the earlier data.
Leveraging the historic information enters through the degree of shrinkage rather than directly through the likelihood.

Quantities such as risk difference and VE are derived as functions of the above parameters.

With only a single trial, we have negligible data to inform the distribution of results on RSVpreF at 360 days.
Therefore, the results are highly dependent on whether the current trial aligns with the historic trial along with the assumptions we make about the trial to trial variability of risk on the RSVpreF arm.

The estimated risk on the RSVpreF arm in both the historic and current trial will be pulled towards one another.
If the historic data for the RSVpreF arm is associated with a lower risk, then the current data will be pulled towards this lower value and vice versa.
In the former case, power will usually be impeded, in the latter case, power will usually increase.

For example, think about a specific case where we look at the risk difference.
Assume that the true $p_0 = 0.1$ but the historic control says $p_h = 0.06$. 
At the same time assume that $p_1 = 0.08$.
With no pooling $rd = p_1 - p_0 = 0.08 - 0.1 = -0.02$ but when we partially pool the control arm, we will end up with something more like $rd = p_1 - p_0 = 0.09 - 0.08 = -0.01$ because the $p_0$ arm has been pulled towards $p_h = 0.06$.
Demonstration below (simulated trials `r n_sim` for scenarios with $p_0$ set as before and $p_1$ set to `r paste0(p_1_sub, collapse = ", ")`.
An informative prior was put on the risk under IG that aligns with a belief that there is a 95% chance that the risk of the outcome is lower than 20%.

```{r}
#| label: historic-ctl-sim-1
#| code-summary: Simulation implementation using historic control 
#| code-fold: true


get_default_sim_hist_rtn <- function(){
  
  d_smry_rd  <- data.table(
    rd_mu = NA_real_, pr_lt_0 = NA_real_, rd_sim = NA_real_
  )
  d_smry_ve  <- data.table(
    ve_mu = NA_real_, pr_gt_20 = NA_real_, ve_sim = NA_real_
  )
  
  return(
    
    list(
      d_smry_rd = d_smry_rd,
      d_smry_ve = d_smry_ve
    )
    
  )
  
}

# only supports a single historic control for the moment because that
# is all we have.
N_ctl = 1500
N_trt = 1500
p_0 = 0.1
p_1 = c(0.075, 0.0875, 0.1, 0.1125, 0.125)
yh = 92
nh = 3495
# n_sim = 10

sim_historic_control <- function(
    N_ctl = 1500, 
    N_trt = 1500,
    p_0 = 0.1,
    p_1 = c(0.075, 0.0875, 0.1, 0.1125, 0.125),
    yh = 92, 
    nh = 3495,
    n_sim = 200
    ){
  
  
  l_res <- list()
  
  j <- 3; i = 3
  for(j in seq_along(p_1)){
    
    y_0 <- rbinom(n_sim, N_ctl, p_0)
    y_1 <- rbinom(n_sim, N_trt, p_1[j])
    
    l_res[[j]] <- pmclapply(1:n_sim, function(i){
      
      ld <- list(
        n0 = N_ctl, y0 = y_0[i], 
        nt = N_trt, yt = y_1[i],
        
        H = 1,
        yh = yh, 
        nh = nh,
        
        # centre at 5% with 2.5 to 10% being well within realms of possibility
        # on the risk scale, converted to log odds scale
        pri_mu = c(qlogis(0.05), 0.25),
        pri_tau_rho = 3, 
      
        # Set the mean to 6% and the pre-existing sample size to 10
        # This is consistent with a 95% probability that the risk on IG at 360 days
        # is less than 20% and a 99% probability that the risk on IG at 360 is less
        # that 32%
        pri_pt = c(alpha(mu, nu), beta(mu, nu))
      )
      
      l_ret <- tryCatch({
        
        outname <- paste0(
          format(Sys.time(), format = "%Y%m%d%H%M%S"),
          "-historic-control-1-", j, "-", i)
        f1 <- m1$sample(
          ld, iter_warmup = 1000, iter_sampling = 2000,
          parallel_chains = 1, chains = 1, refresh = 0, show_exceptions = T,
          max_treedepth = 10, adapt_delta = 0.99,
          output_dir = "/home/mark/Documents/rsv-grant/tmp",
          output_basename = outname)
        
        if(is.null(f1) ){ 
          
          if(is.null(f1)){
            message_parallel("Scenario ", j, " trial ", i, " model issue, f1 is null")  
          }
          
          return(get_default_sim_hist_rtn())
          
        }
        
        # Risk difference 
        d_tmp <- data.table(f1$draws(variables = c("rd"), format = "matrix"))
        d_post <- melt(d_tmp, measure.vars = names(d_tmp))
        d_smry_rd <- d_post[, .(rd_mu = mean(value),
                                pr_lt_0 = mean(value < 0)
                                )]
        d_smry_rd[, rd_sim := p_1[j] - p_0]
        
        d_tmp <- data.table(f1$draws(variables = c("ve"), format = "matrix"))
        d_post <- melt(d_tmp, measure.vars = names(d_tmp))
        d_smry_ve <- d_post[, .(ve_mu = mean(value),
                                pr_gt_20 = mean(value > 0.2)
                                )]
        d_smry_ve[, ve_sim := (1 - (p_1[j]/p_0))]
        
        list(
          d_smry_rd = d_smry_rd,
          d_smry_ve = d_smry_ve
        )
        
        
      }, warning = function(war) {
        
        message_parallel("Scenario ", j, " trial ", i, " warn ", war)  
        return(get_default_sim_hist_rtn())
        
      }, error = function(err) {
        
        message_parallel("Scenario ", j, " trial ", i, " err ", err)
        return(get_default_sim_hist_rtn())
        
      }, finally = {
        # noop
      } )
    
      
      
    }, mc.cores = mc_cores, title = paste0("Historic control, p_1 ", p_1[j]))
    
  }
  l_res
}




```

```{r}
#| label: sim-historic-processing
#| code-summary: Results processing
#| code-fold: true


combine_sim_hist_data <- function(
    l_res_sim
){
  
  d_smry_rd <- rbindlist(lapply(1:length(l_res_sim), function(ii) {
    d_smry_rd <- rbindlist(lapply(1:length(l_res_sim[[ii]]), function(jj) {
      # message("Scenario ", ii, " trial ", jj)
      l_res_sim[[ii]][[jj]]$d_smry_rd
    }), idcol = "i_sim")
    d_smry_rd
  }), idcol = "i_scenario")
  
  d_smry_ve <- rbindlist(lapply(1:length(l_res_sim), function(ii) {
    d_smry_ve <- rbindlist(lapply(1:length(l_res_sim[[ii]]), function(jj) {
      l_res_sim[[ii]][[jj]]$d_smry_ve
    }), idcol = "i_sim")
    d_smry_ve
  }), idcol = "i_scenario")
  
  return(
    list(
      d_smry_rd = d_smry_rd,
      d_smry_ve = d_smry_ve
      
    )
  )
  
}





```

```{r}
#| label: historic-sim-1
#| code-summary: Run historic control simulation at N = 3000 total sample size 
#| code-fold: true

N_ctl <- N_trt <- 1500 
  
l_res_sim_1 <- sim_historic_control(
    N_ctl = N_ctl, 
    N_trt = N_trt,
    p_0 = 0.1,
    p_1 = c(0.075, 0.0875, 0.1, 0.1125, 0.125),
    yh = 92, 
    nh = 3495,
    n_sim = 3000
)



l_res_sim_2 <- sim_historic_control(
    N_ctl = N_ctl, 
    N_trt = N_trt,
    p_0 = 0.1,
    p_1 = c(0.075, 0.0875, 0.1, 0.1125, 0.125),
    yh = 350, 
    nh = 3495,
    n_sim = 3000
)
```

@fig-pwr_2-rd suggests there isn't much benefit in using the historic controls via partial pooling. 
An alternative method for the power prior could be adopted to evaluate whether that shows any improvement. 
The alternative simply incorporates the prior data into a scaled version of the likelihood.

```{r}
#| echo: FALSE
#| label: fig-pwr_2-rd
#| fig-cap: 'Power curves by effect size (risk difference) showing impact when historic control is lower and when the historic control is aligned with the present data.'
#| fig.height: 4.5
#| fig.width: 4.5
#| fig-pos: H

l_res_comb_1 <- combine_sim_hist_data(l_res_sim_1)
l_res_comb_2 <- combine_sim_hist_data(l_res_sim_2)

if(any(is.na(l_res_comb_1$d_smry_rd$rd_mu))){
  d_tmp <- l_res_comb_1$d_smry_rd[is.na(rd_mu)]  
  message("Of the ", length(l_res_sim_1[[1]]), ", ", nrow(d_tmp), 
          " had NA values for the posterior summary. These results will be removed.")
  
}

d_fig_2a <- l_res_comb_1$d_smry_rd[!is.na(rd_mu), .(pwr = mean(pr_lt_0 > 0.975)), 
                                 keyby = .(i_scenario, rd_sim)]
d_fig_2b <- l_res_comb_2$d_smry_rd[!is.na(rd_mu), .(pwr = mean(pr_lt_0 > 0.975)), 
                                 keyby = .(i_scenario, rd_sim)]

d_fig_2 <- rbind(
  cbind(data.table(p_h = sprintf("%.1f%%", 100*92/3495)), d_fig_2a),
  cbind(data.table(p_h = sprintf("%.1f%%", 100*175/3495)), d_fig_2b)
)

tit <- sprintf("Power curve (N_trt = %.0f, N_ctl = %.0f)", N_trt, N_ctl)
ggplot(d_fig_1, aes(x = x_rd, y = win_rd)) +
  geom_smooth(se = F, lwd = 0.3, col = 1, 
              method = "gam", formula = y ~ s(x, bs = "ps")) +
  geom_point(data = d_fig_2,
             aes(x = rd_sim, y = pwr, col = p_h),
             inherit.aes = F) +
  scale_y_continuous("Power", limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  scale_x_continuous("Risk difference (p1-p0)", breaks = seq(-0.03, 0.03, by = 0.01)) +
  scale_color_discrete("Risk in historic study for RSVpreF") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  ggtitle(tit)
```




### Alternative approaches

If a set of applicable outcomes could be ranked, then an ordinal outcome might lead to some efficiency and power gains.
For the sake of the simulation, the following categories and distribution are assumed.
The example below is based on a situation where the treatment effect is assumed to be an odds ratio equal to 0.7.

```{r}
#| label: ordinal-setup
#| code-summary: Setup constants for ordinal outcome variable 
#| code-fold: true


# Define categories and probabilities under standard treatment
cats <- c(
  "death", #  - all cause
  "icu", #  - all cause
  "hosp",  
  "MA",  
  "no rsv")


p_soc <- c(0.01, 0.01, 0.03, 0.1, 0.85) 
names(p_soc) <- cats
stopifnot(sum(p_soc) == 1)

compute_probs <- function(p_soc, or_new_trt){
  cum_p_soc <- cumsum(p_soc)
  # Logit transformation of cumulative probabilities for standard treatment
  lo_cum_probs_soc <- qlogis(cum_p_soc[-length(cum_p_soc)])
  # Adjust logit cumulative probabilities for new treatment
  lo_cum_p_new <- lo_cum_probs_soc + log(or_new_trt)
  cum_p_new <- plogis(lo_cum_p_new)
  cum_p_new <- c(cum_p_new, 1)  # Ensure last category is 1
  p_new <- diff(c(0, cum_p_new))  # Convert
  m <- rbind(p_new, p_soc)  
  colnames(m) <- cats
  m
}

round(compute_probs(p_soc, or_new_trt = 0.7), 4)

```

Assume single cross sectional observation for the analysis.
A units' worst state over the 360 days since birth is taken. 
The approach might evaluate treatment effects directly or use the model to estimate probabilities of being at various states.

```{r}
#| label: ordinal-data
#| code-summary: Data generation for ordinal outcome definition 
#| code-fold: true


# only supports a single historic control for the moment because that
# is all we have.
get_ordinal_data <- function(
    N = 3000, 
    or_new_trt = 0.7,
    # death
    # icu
    # hosp 
    # MA 
    # no rsv
    p_soc = c(0.01, 0.01, 0.03, 0.1, 0.85) 
    ){
  
  cum_p_soc <- cumsum(p_soc)  
  # Logit transformation of cumulative probabilities for standard treatment
  lo_cum_probs_soc <- qlogis(cum_p_soc[-length(cum_p_soc)])


  # Adjust logit cumulative probabilities for new treatment
  lo_cum_p_new <- lo_cum_probs_soc + log(or_new_trt)
  cum_p_new <- plogis(lo_cum_p_new)
  cum_p_new <- c(cum_p_new, 1)  # Ensure last category is 1
  p_new <- diff(c(0, cum_p_new))  # Convert back to category probabilities

  # 1:1 allocation
  trt <- rep(c("SoC", "New"), each = N / 2)
  p_mat <- rbind(p_soc, p_new)
    
  outcomes <- unlist(lapply(1:length(trt), function(i) {
    sample(cats, 1, prob = p_mat[(trt[i] == "New") + 1, ])
  }))
    
  data.table(trt, outcome = factor(outcomes, levels = rev(cats)))

}

```

Based on a single dataset, we can see some improvement in precision on the risk estimates.

```{r}
#| label: proportional-odds-vs-logistic
#| code-summary: Proportional odds vs logistic results based on single data set
#| code-fold: true


d_trial <- get_ordinal_data(N = 3000, 0.7)
d_trial[outcome == "MA", y_bin := 1]
d_trial[outcome != "MA", y_bin := 0]
d_trial[trt == "SoC", x := 0]
d_trial[trt == "New", x := 1]

h1 <- glm(y_bin ~ x, data = d_trial, family = "binomial")

d_trial[, outcome := factor(outcome, levels = cats)]
h2 <- MASS::polr(outcome ~ x, data = d_trial, Hess = TRUE)

# trt arm
marginaleffects::predictions(h1, newdata = data.table(x = 0:1), type = "response")
# compare results to those for MA below
marginaleffects::predictions(h2, newdata = data.table(x = 0:1), type = "probs")
```


```{r}
#| label: ordinal-sim-implementation
#| code-summary: Simulation implementation based on ordinal outcome definition
#| code-fold: true

risk_diff_on_po_opt <- function(or_new_trt, rd_target){
  cum_p_soc <- cumsum(p_soc)  
  # Logit transformation of cumulative probabilities for standard treatment
  lo_cum_probs_soc <- qlogis(cum_p_soc[-length(cum_p_soc)])


  # Adjust logit cumulative probabilities for new treatment
  lo_cum_p_new <- lo_cum_probs_soc + log(or_new_trt)
  cum_p_new <- plogis(lo_cum_p_new)
  cum_p_new <- c(cum_p_new, 1)  # Ensure last category is 1
  p_new <- diff(c(0, cum_p_new))  # Convert back to category probabilities
  
  (rd_target - (p_new[4] - p_soc[4]))^2
}


get_default_sim_ord_rtn <- function(){
  d_smry_p  <- CJ(trt = c("SoC", "New"), outcome = cats)
  d_smry_p[trt == "New", x := 1]
  d_smry_p[trt == "SoC", x := 0]
  d_smry_p[, `:=`(
    N = NA_integer_,
    p_sim = NA_real_,
    p_obs = NA_real_,
    p_mu_ord = NA_real_,
    p_mu_bin = NA_real_
  )]
  setcolorder(d_smry_p, c(
    "trt", "x", "outcome", "N", "p_sim", "p_obs", "p_mu_ord", "p_mu_bin"
  ))
  
  d_smry_rd <- data.table(
    outcome = cats, rd_sim = NA_real_, rd_obs = NA_real_, 
    rd_mu_ord = NA_real_, rd_mu_bin = NA_real_, pr_lt_0 = NA_real_
  )
  
  d_smry_b <- data.table(
    model = c("ordinal", "logistic"),
    rd_sim = NA_real_,
    b_sim = NA_real_,
    b_mu = NA_real_,
    q_025 = NA_real_,
    q_975 = NA_real_,
    pr_lt_0 = NA_real_)
  
  return(
    
    list(
      d_smry_p = d_smry_p,
      d_smry_rd = d_smry_rd,
      d_smry_b = d_smry_b
    )
    
  )
  
}


N = 3000
# now the control arm is the distribution across the possible states.
p_soc = c(0.01, 0.01, 0.03, 0.1, 0.85) 
p_1 = c(0.07, 0.08, 0.09, 0.1, 0.12)
n_sim = 200


sim_ordinal <- function(
    N = 3000, # total sample size 
    p_soc = c(0.01, 0.01, 0.03, 0.1, 0.85)  ,
    # scenarios characterising treatment group risk for MA rsv+
    p_1 = c(0.07, 0.08, 0.09, 0.1, 0.12) ,
    n_sim = 200
  ){
  
  # Compute the ORs that are consistent with risk differences that we 
  # have looked at previously.
  rd_sub <- p_1 - p_soc[4]
  or_new_trt <- length(p_1)
  for(i in seq_along(p_1)){
  
    or_new_trt[i] <- optimize(
      risk_diff_on_po_opt, 
      c(0, 10), tol = 0.0001, 
      rd_target = p_1[i] - p_soc[4])$minimum
  
  }

  d_grid <- CJ(trt = c("SoC", "New"), outcome = cats)
  d_grid[, outcome := factor(outcome, levels = cats)]
  
  l_res <- list()
  
  i <- j <- 3
  for(j in seq_along(or_new_trt)){
    
    l_res[[j]] <- pmclapply(1:n_sim, function(i){
      
      d_trial <- get_ordinal_data(N = N, or_new_trt[j])
      
      d_trial[trt == "New", x := 1]
      d_trial[trt == "SoC", x := 0]
      d_trial <- d_trial[order(x)]
      
      d_trial_grp <- d_trial[, .N, keyby = .(trt, x, outcome)]
      # Accounting for chance that some groups have zero counts
      d_trial_grp <- merge(d_trial_grp, d_grid, by = c("trt", "outcome"), all.y = T)
      d_trial_grp[is.na(x) & trt == "New", x := 1]
      d_trial_grp[is.na(x) & trt == "SoC", x := 0]
      d_trial_grp[is.na(N), N := 0]
      
      d_trial_grp[, p_obs := N / (nrow(d_trial)/2)]
      d_trial_grp <- d_trial_grp[order(x)]
      
      d_trial_bin <- copy(d_trial_grp)
      d_trial_bin[outcome == "MA", evt := 1]
      d_trial_bin[outcome != "MA", evt := 0]
      d_trial_bin <- d_trial_bin[, .(N = sum(N)), keyby = .(trt, evt)]
      d_trial_bin <- dcast(d_trial_bin, trt ~ evt, value.var = "N")
      setnames(d_trial_bin, "0", "failure")
      setnames(d_trial_bin, "1", "success")
      d_trial_bin[, n := failure + success]
      d_trial_bin[trt == "New", x := 1]
      d_trial_bin[trt == "SoC", x := 0]
      d_trial_bin <- d_trial_bin[order(x)]
      d_trial_bin[, p_obs := success/n]
      
      
      
      l_ret <- tryCatch({
        
        ld <- list(
          N = nrow(d_trial_grp),
          K = length(cats),
          P = 1,
          y = as.integer(d_trial_grp$outcome),
          X = as.matrix(d_trial_grp$x, ncol = 1, drop = F ),
          wgt = d_trial_grp$N,
          pri_b_s = 3, pri_cuts_s = 3
        )
        
        outname <- paste0(
          format(Sys.time(), format = "%Y%m%d%H%M%S"),
          "-ordinal-02-", j, "-", i)
        
        f4 <- m3$sample(
          ld, iter_warmup = 1000, iter_sampling = 2000,
          parallel_chains = 1, chains = 1, refresh = 0, show_exceptions = F,
          max_treedepth = 10, adapt_delta = 0.96,
          output_dir = "/home/mark/Documents/rsv-grant/tmp",
          output_basename = outname)
      
        ld <- list(
          N = nrow(d_trial_bin),
          y = d_trial_bin$success, n = d_trial_bin$n,
          X = as.matrix(d_trial_bin$x, ncol = 1, drop = F )
        )
        
        outname <- paste0(
          format(Sys.time(), format = "%Y%m%d%H%M%S"),
          "-logistic-", j, "-", i)
        
        f5 <- m4$sample(
          ld, iter_warmup = 1000, iter_sampling = 2000,
          parallel_chains = 1, chains = 1, refresh = 0, show_exceptions = T,
          max_treedepth = 10, adapt_delta = 0.95,
          output_dir = "/home/mark/Documents/rsv-grant/tmp",
          output_basename = outname)
      
        if(is.null(f4) || is.null(f5) ){ 
          
          if(is.null(f4)){
            message_parallel("Scenario ", j, " trial ", i, " model issue, f4 is null")  
          }
          if(is.null(f5)){
            message_parallel("Scenario ", j, " trial ", i, " model issue, f5 is null")  
          }
          
          return(get_default_sim_ord_rtn())
          
        }
        
        # Risks by outcome from ordinal
        d_tmp <- data.table(f4$draws(variables = c("p0", "p1"), format = "matrix"))
        d_post <- melt(d_tmp, measure.vars = names(d_tmp))
        d_post[, ix := gsub("p.\\[", "", variable, fixed = F)]
        d_post[, ix := as.numeric(gsub("\\]", "", ix, fixed = F))]
        d_post[variable %like% "p0", trt := "SoC"]
        d_post[variable %like% "p1", trt := "New"]
        d_smry_ord <- d_post[, .(p_mu_ord = mean(value)), keyby = .(trt, ix)]
        d_smry_ord[, outcome := cats[length(cats) + 1 -ix]]
        d_smry_ord[, ix := NULL]
        d_smry_ord <- merge(d_smry_ord, d_trial_grp, by = c("trt", "outcome"), all = T)
        d_smry_ord[, outcome := factor(outcome, levels = cats)]
        d_smry_ord <- d_smry_ord[order(x, outcome)]
        d_smry_ord[trt == "SoC", p_sim := p_soc]
        m_tmp <- compute_probs(p_soc, or_new_trt[j])
        d_smry_ord[trt == "New", p_sim := m_tmp[rownames(m_tmp) == "p_new", ]]
      
        # Risks by outcome from logistic based on aggregated data
        d_tmp <- data.table(f5$draws(variables = c("p0", "p1"), format = "matrix"))
        d_post <- melt(d_tmp, measure.vars = names(d_tmp))
        d_post[variable %like% "p0", trt := "SoC"]
        d_post[variable %like% "p1", trt := "New"]
        d_smry_bin <- d_post[, .(p_mu_bin = mean(value)), keyby = .(trt)]
        d_smry_bin[, outcome := "MA"]
        
        d_smry_p <- merge(d_smry_ord, d_smry_bin, by = c("trt", "outcome"), all.x = T)
        setcolorder(d_smry_p, c("trt", "x", "outcome", "N", "p_sim", "p_obs", "p_mu_ord", "p_mu_bin"))
        d_smry_p[, outcome := factor(outcome, levels = cats)]
        d_smry_p <- d_smry_p[order(outcome, x)]
      
        
        # Risk difference 
        d_tmp <- data.table(f4$draws(variables = c("rd"), format = "matrix"))
        d_post <- melt(d_tmp, measure.vars = names(d_tmp))
        d_post[, ix := gsub("rd\\[", "", variable, fixed = F)]
        d_post[, ix := as.numeric(gsub("\\]", "", ix, fixed = F))]
        d_smry_ord <- d_post[, .(rd_mu_ord = mean(value),
                                 pr_lt_0 = mean(value < 0)
                                 ), keyby = .(ix)]
        d_smry_ord[, outcome := cats[length(cats) + 1 -ix]]
        d_smry_ord[, ix := NULL]
        d_smry_ord <- merge(
          d_smry_ord, 
          dcast(d_trial_grp, outcome ~ trt, value.var = "p_obs")[
            , .(outcome, rd_obs = New - SoC)], 
          by = c("outcome"), all = T)
        d_smry_ord[, outcome := factor(outcome, levels = cats)]
        
        d_tmp <- data.table(f5$draws(variables = c("rd"), format = "matrix"))
        d_post <- melt(d_tmp, measure.vars = names(d_tmp))
        d_smry_bin <- d_post[, .(rd_mu_bin = mean(value))]
        d_smry_bin[, outcome := factor("MA", levels = cats)]
        
        d_smry_rd <- merge(d_smry_ord, d_smry_bin, by = "outcome", all.x = T)
        d_smry_rd[, rd_sim := rd_sub[j]]
        
        setcolorder(d_smry_rd, c("outcome", "rd_sim", "rd_obs", "rd_mu_ord", "rd_mu_bin"))
        d_smry_rd[, outcome := factor(outcome, levels = cats)]
        d_smry_rd <- d_smry_rd[order(outcome)]
        
        # Treatment effect
        d_tmp <- data.table(f4$draws(variables = c("b"), format = "matrix"))
        d_post <- melt(d_tmp, measure.vars = names(d_tmp))
        d_smry_ord_b <- d_post[, .(b_mu = mean(value),
                               q_025 = quantile(value, prob = 0.025),
                               q_975 = quantile(value, prob = 0.975),
                               pr_lt_0 = mean(value < 0))]
        d_smry_ord_b[, b_sim := round(log(or_new_trt[j]), 5)]
        d_smry_ord_b[, rd_sim := rd_sub[j]]
        
              
        d_tmp <- data.table(f5$draws(variables = c("b"), format = "matrix"))
        d_post <- melt(d_tmp, measure.vars = names(d_tmp))
        d_smry_bin_b <- d_post[, .(b_mu = mean(value),
                               q_025 = quantile(value, prob = 0.025),
                               q_975 = quantile(value, prob = 0.975),
                               pr_lt_0 = mean(value < 0))]
        d_smry_bin_b[, b_sim := NA_real_]
        d_smry_bin_b[, rd_sim := rd_sub[j]]
        
        d_smry_b <- rbind(
          cbind(model = "ordinal", d_smry_ord_b),
          cbind(model = "logistic", d_smry_bin_b),
          fill = T
        )
        setcolorder(d_smry_b, c("model", "rd_sim", "b_sim"))
        
        
        list(
          d_smry_p = d_smry_p,
          d_smry_rd = d_smry_rd,
          d_smry_b = d_smry_b
        )
        
      }, warning = function(war) {
        
        message_parallel("Scenario ", j, " trial ", i, " warn ", war)  
        return(get_default_sim_ord_rtn())
        
      }, error = function(err) {
        
        message_parallel("Scenario ", j, " trial ", i, " err ", err)
        return(get_default_sim_ord_rtn())
        
      }, finally = {
        # noop
      } )
      
      
      
    }, mc.cores = mc_cores, title = paste0("Ordinal, OR ", or_new_trt[j]))
    
  }
  
  l_res
}

```


```{r}
#| label: ordinal-sim-processing
#| code-summary: Process results from simulation for ordinal outcome 
#| code-fold: true


combine_sim_ord_data <- function(
    l_res_sim
){
  
  d_smry_p <- rbindlist(lapply(1:length(l_res_sim), function(ii) {
    d_smry_p <- rbindlist(lapply(1:length(l_res_sim[[ii]]), function(jj) {
      # message("Scenario ", ii, " trial ", jj)
      l_res_sim[[ii]][[jj]]$d_smry_p
    }), idcol = "i_sim")
    d_smry_p
  }), idcol = "i_scenario")
  
  d_smry_rd <- rbindlist(lapply(1:length(l_res_sim), function(ii) {
    d_smry_rd <- rbindlist(lapply(1:length(l_res_sim[[ii]]), function(jj) {
      l_res_sim[[ii]][[jj]]$d_smry_rd
    }), idcol = "i_sim")
    d_smry_rd
  }), idcol = "i_scenario")
  
  d_smry_b <- rbindlist(lapply(1:length(l_res_sim), function(ii) {
    d_smry_b <- rbindlist(lapply(1:length(l_res_sim[[ii]]), function(jj) {
      l_res_sim[[ii]][[jj]]$d_smry_b
    }), idcol = "i_sim")
    d_smry_b
  }), idcol = "i_scenario")
  
  return(
    list(
      d_smry_p = d_smry_p,
      d_smry_rd = d_smry_rd,
      d_smry_b = d_smry_b
      
    )
  )
  
}


```


```{r}
#| label: ordinal-sim-1
#| code-summary: Run simulation at original N = 3000 sample size 
#| code-fold: true

N <- 3000
l_res_sim_1 <- sim_ordinal(
  N = N, # total sample size 
  p_soc = c(0.01, 0.01, 0.03, 0.1, 0.85)  ,
  # scenarios characterising treatment group risk for MA rsv+
  p_1 = c(0.07, 0.075, 0.08, 0.085, 0.09, 0.1, 0.12) ,
  n_sim = 3000
)
```

@fig-pwr-3 suggests there is some benefit in running the proportional odds model on the complete data rather than a logistic regression on the dichotomised scale.

```{r}
#| echo: FALSE
#| label: fig-pwr-3
#| fig-cap: 'Power curves (risk difference) showing logistic vs ordinal at original sample size.'
#| fig.height: 4.5
#| fig.width: 4.5
#| fig-pos: H

l_res_comb_1 <- combine_sim_ord_data(l_res_sim_1)

if(any(is.na(l_res_comb_1$d_smry_b$b_mu))){
  d_tmp <- l_res_comb_1$d_smry_b[is.na(b_mu)]  
  message("Of the ", nrow(l_res_comb_1$d_smry_b), " simulation results, ", nrow(d_tmp), 
          " had NA values for the posterior summary for b_mu. These results will be removed.")
  
}

d_fig_3 <- l_res_comb_1$d_smry_b[!is.na(b_mu), .(pwr = mean(pr_lt_0 > 0.975)), 
                                 keyby = .(i_scenario, model, rd_sim)]


tit <- sprintf("Power curve (N_trt = %.0f, N_ctl = %.0f)", N/2, N/2)
ggplot(d_fig_1, aes(x = x_rd, y = win_rd)) +
  geom_smooth(se = F, lwd = 0.3, col = 1, 
              method = "gam", formula = y ~ s(x, bs = "ps")) +
  geom_line(data = d_fig_3,
             aes(x = rd_sim, y = pwr, col = model, group = model),
             inherit.aes = F, lwd = 0.3, lty = 2) +
  geom_point(data = d_fig_3,
             aes(x = rd_sim, y = pwr, col = model),
             inherit.aes = F) +
  scale_y_continuous("Power", limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  scale_x_continuous("Risk difference (p1-p0)", breaks = seq(-0.03, 0.03, by = 0.01)) +
  scale_color_discrete("Risk in historic study for RSVpreF") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  ggtitle(tit)
```

```{r}
#| label: ordinal-sim-2
#| code-summary: Run simulation at original N = 2000 sample size 
#| code-fold: true


N = 2000
l_res_sim_2 <- sim_ordinal(
  N = N, # total sample size 
  p_soc = c(0.01, 0.01, 0.03, 0.1, 0.85)  ,
  # scenarios characterising treatment group risk for MA rsv+
  p_1 = c(0.07, 0.075, 0.08, 0.085, 0.09, 0.1, 0.12) ,
  n_sim = 3000
)
```

@fig-pwr-4 suggests that reducing the sample size to 2000 will still produce operating characteristics that are above the original approach.

```{r}
#| echo: FALSE
#| label: fig-pwr-4
#| fig-cap: 'Power curves (risk difference) showing logistic vs ordinal at lower sample size.'
#| fig.height: 4.5
#| fig.width: 4.5
#| fig-pos: H

l_res_comb_2 <- combine_sim_ord_data(l_res_sim_2)

if(any(is.na(l_res_comb_2$d_smry_b$b_mu))){
  d_tmp <- l_res_comb_2$d_smry_b[is.na(b_mu)]  
  message("Of the ", nrow(l_res_comb_2$d_smry_b), " simulation results, ", nrow(d_tmp), 
          " had NA values for the posterior summary for b_mu. These results will be removed.")
  
}

d_fig_4 <- l_res_comb_2$d_smry_b[!is.na(b_mu), .(pwr = mean(pr_lt_0 > 0.975)), 
                                 keyby = .(i_scenario, model, rd_sim)]


tit <- sprintf("Power curve (N_trt = %.0f, N_ctl = %.0f)", N/2, N/2)
ggplot(d_fig_1, aes(x = x_rd, y = win_rd)) +
  geom_smooth(se = F, lwd = 0.3, col = 1, 
              method = "gam", formula = y ~ s(x, bs = "ps")) +
  geom_line(data = d_fig_4,
             aes(x = rd_sim, y = pwr, col = model, group = model),
             inherit.aes = F, lwd = 0.3, lty = 2) +
  geom_point(data = d_fig_4,
             aes(x = rd_sim, y = pwr, col = model),
             inherit.aes = F) +
  scale_y_continuous("Power", limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  scale_x_continuous("Risk difference (p1-p0)", breaks = seq(-0.03, 0.03, by = 0.01)) +
  scale_color_discrete("Risk in historic study for RSVpreF") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  ggtitle(tit)
```


```{r}
#| label: ordinal-sim-3
#| code-summary: Run simulation at original N = 1000 sample size 
#| code-fold: true


N = 1000
l_res_sim_3 <- sim_ordinal(
  N = N, # total sample size 
  p_soc = c(0.01, 0.01, 0.03, 0.1, 0.85)  ,
  # scenarios characterising treatment group risk for MA rsv+
  p_1 = c(0.07, 0.075, 0.08, 0.085, 0.09, 0.1, 0.12) ,
  n_sim = 3000
)
```

@fig-pwr-5 shows the results at a total sample size of 1000.

```{r}
#| echo: FALSE
#| label: fig-pwr-5
#| fig-cap: 'Power curves (risk difference) showing logistic vs ordinal at lower sample size.'
#| fig.height: 4.5
#| fig.width: 4.5
#| fig-pos: H

l_res_comb_3 <- combine_sim_ord_data(l_res_sim_3)

if(any(is.na(l_res_comb_3$d_smry_b$b_mu))){
  d_tmp <- l_res_comb_3$d_smry_b[is.na(b_mu)]  
  message("Of the ", nrow(l_res_comb_3$d_smry_b), " simulation results, ", nrow(d_tmp), 
          " had NA values for the posterior summary for b_mu. These results will be removed.")
  
}

d_fig_5 <- l_res_comb_3$d_smry_b[!is.na(b_mu), .(pwr = mean(pr_lt_0 > 0.975)), 
                                 keyby = .(i_scenario, model, rd_sim)]


tit <- sprintf("Power curve (N_trt = %.0f, N_ctl = %.0f)", N/2, N/2)
ggplot(d_fig_1, aes(x = x_rd, y = win_rd)) +
  geom_smooth(se = F, lwd = 0.3, col = 1, 
              method = "gam", formula = y ~ s(x, bs = "ps")) +
  geom_line(data = d_fig_5,
             aes(x = rd_sim, y = pwr, col = model, group = model),
             inherit.aes = F, lwd = 0.3, lty = 2) +
  geom_point(data = d_fig_5,
             aes(x = rd_sim, y = pwr, col = model),
             inherit.aes = F) +
  scale_y_continuous("Power", limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  scale_x_continuous("Risk difference (p1-p0)", breaks = seq(-0.03, 0.03, by = 0.01)) +
  scale_color_discrete("Risk in historic study for RSVpreF") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  ggtitle(tit)
```


```{r}
#| label: tidy-up
#| code-summary: Shut down timer
#| code-fold: true

toc(log = TRUE, quiet = TRUE)
```

Simulations run: `r unlist(tic.log(format = TRUE))`.

<!-- ::: {#refs} -->
<!-- ::: -->





