---
title: "Simulation - RSV study evaluating RSVpreF vs IG"
subtitle: "Sequential analyses"
author: "maj"
date: "2025-04-03"
date-modified: last-modified
bibliography: refs.bib
csl: biomed-central.csl
embed-resources: true
---

```{r}  
#| label: setup
#| code-summary: Libraries and globals
#| code-fold: true

suppressWarnings(suppressPackageStartupMessages(library(cmdstanr)))
library(ggplot2)
library(parallel)
library(mcprogress)
suppressWarnings(suppressPackageStartupMessages(library(tictoc)))
library(poisson)
suppressWarnings(suppressPackageStartupMessages(library(data.table)))
library(survival)
library(flexsurv)
suppressWarnings(suppressPackageStartupMessages(library(extraDistr)))

suppressPackageStartupMessages(suppressWarnings(library("gt")))

tic()

set.seed(92818785)

mc_cores <- 50

tok <- unlist(tstrsplit(getwd(), split = "/"))
if(tok[length(tok)] == "rsv-grant"){
  mod_01 <- cmdstanr::cmdstan_model("./stan/log-logistic-aft-01.stan")
  
  output_dir_res <- "./out"
} else {
  mod_01 <- cmdstanr::cmdstan_model("../stan/log-logistic-aft-01.stan")
  output_dir_res <- "../out"
}


```

## Background

### Aim

To compare effectiveness of IG vs RSVpreF for protection against RSV in the first 360 days of life.

### Interventions

+ IG - monoclonal antibody given to infant at birth
+ RSVpreF - maternal vaccine given after 32 weeks gestation

### Randomisation

Balanced 1:1 randomisation. 
Variations from this might be possible if historical controls can be used.

### Primary outcome

Options:

+ binary outcome variable indicating medically attended per the definition of MATISSE study:

  + MA-RTI visit AND RSV-positive test result AND one or more of the following
    + Fast breathing (RR $\ge 60$  bpm for $<2$  months of age [$<60$ days of age], $\ge 50$ bpm for $2 - <12$ months of age, or $\ge 40$ bpm for 12–24 months of age)
    + SpO2 $<95%$
    + Chest wall indrawing
    
+ time to first occurrence of the above (in infants only) from birth to day 360

### Estimand

Primary: population perspective, i.e. the estimated comparative effectiveness of each implemented strategy with adherence as observed under real-world conditions.

### Follow up

Follow up to age 360 days for both groups.

### Prior information

Weakly regularising priors are adopted as soft-constraints on all model parameters.

### Resource Constraints

Sample size up to approximately 3000 dyads.

## Methods

Here we consider the use of interim analysis assessing futility and superiority as stopping criteria followed by a final analysis at the maximum sample size.
Variations on this adaptation might consider sample size re-estimation.

An AFT survival model (log-logistic) is used to model the time to first occurrence of MA-RSV, censoring at day 360.
In contrast to a CoxPH model, an AFT estimates how different factors speed up or slow down the time to event.
Instead of focusing on the instantaneous risk (hazard) of the event happening at any given time (as in the Cox model), the AFT model directly models survival time.

Covariates are incorporated as a linear predictor for the scale parameter on the log scale with exponentiated terms interpreted as acceleration factors.

+ AF > 1: The event takes longer to occur (slower time to failure implying a protective effect).
+ AF < 1: The event happens more quickly (shorter time to failure implying harmful effect).
+ AF = 1: No effect; time to event is the same across groups.

The RMST is the average amount of time that patients survive (free of the event) up to a specific point in time (here day 360).
In contrast to the median survival time (the time by which half the units have had the event) which captures a point in time, the RMST considers the full survival distribution (up to the time of interest).
For example, if the difference in the RMST is 20 days in the treatment group, it tells us that, on average, units that received the treatment were event free for 20 days longer than those that received the control.

RMST can be estimated by integrating the survival curves to day 360 and then the difference in RMST across groups can be used to evaluate treatment differences, although this will be equivalent to simply comparing the treatment effects when no clinically important difference is defined.

Within this simulation, the probability thresholds for superiority and futilty were:

+ Superiority 0.99, 0.98, 0.98, 0.9725 from first to last analysis respectively with the last analysis undertaken once all participants have completed 360 days follow up and 
+ Futility 0.35 at all analyses

These were used for comparing the difference in RMST relative to a zero reference point; superiority looking for extended RMST in the treatment group and futility looking for reduced RMST.

Use of additional covariates in the model needs to be investigated, but is, in principle possible.

Provisional results based on a classical parametric AFT survival model (log-logistic) and a normal approximation for the joint posterior.

## Operating characteristics

### Reference design

A reference design with single analysis based on a logistic regression at 3000 dyads has been considered previously.

In order to develop a sequential design, it is important to consider how to handle those participants that have not completed follow up.
Ignoring these units would cause bias in the treatment effects.
Some form of imputation would be the standard way address the missingness.
The alternative approach is to adopt a survival perspective as is done here.

### Sequential design

```{r}
#| label: load-results-N3000
#| code-summary: Result set
#| code-fold: true

l_res <- qs::qread(file.path(output_dir_res, "20250403171118-sim-res.qs"))


d_sim_dat <- rbindlist(lapply(1:length(l_res), function(i){
  l_res[[i]]$d
}), idcol = "id_sim")

d_sim_res <- rbindlist(lapply(1:length(l_res), function(i){
  l_res[[i]]$d_pars
}), idcol = "id_sim")

d_sim_res[, sup := fifelse(pr_drmst_gt_0 > p_sup, 1, 0)]
d_sim_res[, fut := fifelse(pr_drmst_gt_0 < p_fut, 1, 0)]
# 
# d_sim_res[id_analy == 2, N_enrol := 2500]
# d_sim_res[id_analy == 3, N_enrol := 3000]

# The static imputation allows us to get unbiased summaries over all interim
# analyses. Because of the potential for early stopping, later interim analyses
# do not reflect a random selection and therefore the summary results would be
# biased.
d_sim_res[, sup := nafill(sup, type = "locf"), keyby = id_sim]
d_sim_res[, fut := nafill(fut, type = "locf"), keyby = id_sim]
d_sim_res[, mu_rmst0 := nafill(mu_rmst0, type = "locf"), keyby = id_sim]
d_sim_res[, mu_rmst1 := nafill(mu_rmst1, type = "locf"), keyby = id_sim]
d_sim_res[, mu_drmst := nafill(mu_drmst, type = "locf"), keyby = id_sim]
d_sim_res[, p0_360 := nafill(p0_360, type = "locf"), keyby = id_sim]
d_sim_res[, p1_360 := nafill(p1_360, type = "locf"), keyby = id_sim]
d_sim_res[, rd_360 := nafill(rd_360, type = "locf"), keyby = id_sim]
d_sim_res[, p0_sim := nafill(p0_sim, type = "locf"), keyby = id_sim]
d_sim_res[, p1_sim := nafill(p1_sim, type = "locf"), keyby = id_sim]
d_sim_res[, rd_sim := p1_sim - p0_sim]
d_sim_res[, scale0 := nafill(scale0, type = "locf"), keyby = id_sim]
d_sim_res[, scale1 := nafill(scale1, type = "locf"), keyby = id_sim]
d_sim_res[, shape := nafill(shape, type = "locf"), keyby = id_sim]

p0 <- sort(unique(d_sim_res$p0_sim))
stopifnot(length(p0) == 1)
p1_uniq <- sort(unique(d_sim_res$p1_sim))



N_tick <- sort(unique(d_sim_res$N_enrol))

N_sims <- d_sim_res[id_analy == 1, .N, keyby = .(p1_sim)][, unique(N)]
```


@fig-pwr_1 shows the cumulative probability of superiority/futility by interim and by risk difference at day 360 with a maximum sample size of 3000 and two interim analyses (number of simulations `r N_sims`).

```{r}
#| echo: FALSE
#| label: fig-pwr_1
#| fig-cap: 'Cumulative probability of decision'
#| fig.height: 6
#| fig.width: 7
#| fig-pos: H
#| lightbox: true

d_fig <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, rd_sim, sup, fut)]

# d_fig[, mean(sup), keyby = .(N, rd_sim)]

d_fig[, rd_sim := factor(rd_sim, levels = p1_uniq - p0)]

d_fig <- rbind(
  cbind(decision = "Superiority", d_fig[, .(pr_dec = mean(sup)), keyby = .(N, rd_sim)]),
  cbind(decision = "Futility", d_fig[, .(pr_dec = mean(fut)), keyby = .(N, rd_sim)])
)
d_fig[, decision := factor(decision, levels = c("Superiority", "Futility"))]

ggplot(
  d_fig, 
  aes(x = N, y = pr_dec, group = rd_sim, col = rd_sim)) +
  geom_line() +
  geom_point(size = 0.7) +
  scale_color_discrete("RD (day 360)") +
  scale_x_continuous("", breaks = N_tick) +
  scale_y_continuous(
    "Cumulative probability of decision", limits = c(0, 1),
    breaks = seq(0, 1, by = 0.1)) +
  labs(caption = "Note: Interim use the available data, final assumes all units followed to 360 days") +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
    plot.caption = element_text(hjust = 0)
  ) +
    guides(col=guide_legend(ncol=1)) +
  facet_wrap(~decision, labeller = label_both) +
  ggtitle("Cumulative probability of decision by enrolment size:")




```


```{r}
#| echo: FALSE
#| label: tbl-pwr1
#| tbl-pos: H
#| tbl-cap: "Cumulative probability of decision by decision type, enrolments and difference in cumulative incidence at day 360"


d_tbl <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, rd_sim, sup, fut)]

d_tbl_sup <- dcast(
  d_tbl[, .(pr_decision = mean(sup)), keyby = .(N, rd_sim)],
  N ~ rd_sim, value.var = "pr_decision"
)

d_tbl <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, rd_sim, sup, fut)]
d_tbl_fut <- dcast(
  d_tbl[, .(pr_decision = mean(fut)), keyby = .(N, rd_sim)],
  N ~ rd_sim, value.var = "pr_decision"
)

d_tbl <- rbind(
  cbind(descision = "Superiority", d_tbl_sup),
  cbind(descision = "Futility", d_tbl_fut)
)

gt(d_tbl) |>
  tab_options(
    table.font.size = "80%"
  ) 
```


@fig-par-est1 shows the distribution of the posterior means from each simulation by analysis and risk difference at day 360.

```{r}
#| echo: FALSE
#| label: fig-par-est1
#| fig-cap: 'Distribution of posterior means'
#| fig.height: 6
#| fig.width: 6
#| fig-pos: H
#| lightbox: true


d_fig <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, rd_sim, rd_360, sup, fut)]
d_fig <- melt(
  d_fig[
    , .(N, rd_sim, rd_360)], 
  measure.vars = c("rd_360"),
  value.name = "mu_rd")
d_fig[, rd_sim := factor(rd_sim, levels = p1_uniq - p0)]

d_fig <- d_fig[, .(
  mu_rd = mean(mu_rd),
  q_025 = quantile(mu_rd, prob = 0.025),
  q_975 = quantile(mu_rd, prob = 0.975)
), keyby = .(N, rd_sim)]

ggplot(
  d_fig, 
  aes(x = N, y = mu_rd)) +
  geom_linerange(aes(ymin = q_025, ymax = q_975), lwd = 0.3) +
  geom_point(size = 0.7) +
  scale_x_continuous("", c(2000, 2500, 3000)) +
  scale_y_continuous(
    "Distribution of mean RD", 
    limits = c(-0.08, 0.05), breaks = seq(-0.08, 0.05, by = 0.02)) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)
  ) +
  guides(fill=guide_legend(nrow=4)) +
  facet_wrap(~rd_sim, labeller = label_both)+
  ggtitle("Distribution of posterior means RD(360) by interim:")




```

@fig-rmst shows the distribution of the posterior means for RMST across the simulations by analysis and risk difference at day 360.

```{r}
#| echo: FALSE
#| label: fig-rmst
#| fig-cap: 'RMST distribution'
#| fig.height: 6
#| fig.width: 6
#| fig-pos: H
#| lightbox: true




# distribution of rmst by risk diff at day 360 and analysis
d_fig <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, rd_sim, mu_rmst0, mu_rmst1, sup, fut)]
d_fig <- melt(
  d_fig[
    , .(N, rd_sim, mu_rmst0, mu_rmst1)], 
  measure.vars = c("mu_rmst0", "mu_rmst1"))
d_fig[, trt := factor(fifelse(variable == "mu_rmst0", 0, 1))]
d_fig[, rd_sim := factor(rd_sim, levels = p1_uniq - p0)]

ggplot(
  d_fig, 
  aes(x = value, group = trt, col = trt)) +
  geom_density() +
  geom_vline(
    data = d_fig[, .(value = mean(value)), keyby = .(N, rd_sim, trt)],
    aes(xintercept = value, group = trt, col = trt),
    lwd = 0.3
  ) +
  scale_color_discrete("Treatment") +
  scale_x_continuous("RMST") +
  # scale_y_continuous(
  #   "Distribution of mean RD", 
  #   limits = c(-0.1, 0.1), breaks = seq(-0.1, 0.1, by = 0.02)) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)
  ) +
  # guides(fill=guide_legend(nrow=4)) +
  facet_grid(rd_sim ~ N, labeller = label_both) +
  ggtitle("RMST(360) by treatment group and RD(360):")

```




```{r}
#| label: load-results-N2500
#| code-summary: Result set
#| code-fold: true

l_res <- qs::qread(file.path(output_dir_res, "20250403171938-sim-res.qs"))


d_sim_dat <- rbindlist(lapply(1:length(l_res), function(i){
  l_res[[i]]$d
}), idcol = "id_sim")

d_sim_res <- rbindlist(lapply(1:length(l_res), function(i){
  l_res[[i]]$d_pars
}), idcol = "id_sim")

d_sim_res[, sup := fifelse(pr_drmst_gt_0 > p_sup, 1, 0)]
d_sim_res[, fut := fifelse(pr_drmst_gt_0 < p_fut, 1, 0)]
# 
# d_sim_res[id_analy == 2, N_enrol := 2500]
# d_sim_res[id_analy == 3, N_enrol := 3000]

# The static imputation allows us to get unbiased summaries over all interim
# analyses. Because of the potential for early stopping, later interim analyses
# do not reflect a random selection and therefore the summary results would be
# biased.
d_sim_res[, sup := nafill(sup, type = "locf"), keyby = id_sim]
d_sim_res[, fut := nafill(fut, type = "locf"), keyby = id_sim]
d_sim_res[, mu_rmst0 := nafill(mu_rmst0, type = "locf"), keyby = id_sim]
d_sim_res[, mu_rmst1 := nafill(mu_rmst1, type = "locf"), keyby = id_sim]
d_sim_res[, mu_drmst := nafill(mu_drmst, type = "locf"), keyby = id_sim]
d_sim_res[, p0_360 := nafill(p0_360, type = "locf"), keyby = id_sim]
d_sim_res[, p1_360 := nafill(p1_360, type = "locf"), keyby = id_sim]
d_sim_res[, rd_360 := nafill(rd_360, type = "locf"), keyby = id_sim]
d_sim_res[, p0_sim := nafill(p0_sim, type = "locf"), keyby = id_sim]
d_sim_res[, p1_sim := nafill(p1_sim, type = "locf"), keyby = id_sim]
d_sim_res[, rd_sim := p1_sim - p0_sim]


p0 <- sort(unique(d_sim_res$p0_sim))
stopifnot(length(p0) == 1)
p1_uniq <- sort(unique(d_sim_res$p1_sim))


N_tick <- sort(unique(d_sim_res$N_enrol))
N_sims <- d_sim_res[id_analy == 1, .N, keyby = .(p1_sim)][, unique(N)]
```


@fig-pwr_2 shows the cumulative probability of superiority/futility by interim and by risk difference at day 360 with a maximum sample size of 2500 and two interim analyses (number of simulations `r N_sims`).

```{r}
#| echo: FALSE
#| label: fig-pwr_2
#| fig-cap: 'Cumulative probability of decision (N = 2800 maximum sample size)'
#| fig.height: 6
#| fig.width: 7
#| fig-pos: H
#| lightbox: true

d_fig <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, rd_sim, sup, fut)]

# d_fig[, mean(sup), keyby = .(N, rd_sim)]

d_fig[, rd_sim := factor(rd_sim, levels = p1_uniq - p0)]

d_fig <- rbind(
  cbind(decision = "Superiority", d_fig[, .(pr_dec = mean(sup)), keyby = .(N, rd_sim)]),
  cbind(decision = "Futility", d_fig[, .(pr_dec = mean(fut)), keyby = .(N, rd_sim)])
)
d_fig[, decision := factor(decision, levels = c("Superiority", "Futility"))]

ggplot(
  d_fig, 
  aes(x = N, y = pr_dec, group = rd_sim, col = rd_sim)) +
  geom_line() +
  geom_point(size = 0.7) +
  scale_color_discrete("RD (day 360)") +
  scale_x_continuous("", breaks = N_tick) +
  scale_y_continuous(
    "Cumulative probability of decision", limits = c(0, 1),
    breaks = seq(0, 1, by = 0.1)) +
  labs(caption = "Note: Interim use the available data, final assumes all units followed to 360 days") +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
    plot.caption = element_text(hjust = 0)
  ) +
    guides(col=guide_legend(ncol=1)) +
  facet_wrap(~decision, labeller = label_both) +
  ggtitle("Cumulative probability of decision by enrolment size:")




```



```{r}
#| echo: FALSE
#| label: tbl-pwr2
#| tbl-pos: H
#| tbl-cap: "Cumulative probability of decision by decision type, enrolments and difference in cumulative incidence at day 360"


d_tbl <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, rd_sim, sup, fut)]

d_tbl_sup <- dcast(
  d_tbl[, .(pr_decision = mean(sup)), keyby = .(N, rd_sim)],
  N ~ rd_sim, value.var = "pr_decision"
)

d_tbl <- d_sim_res[, .(id_sim, id_analy, N = N_enrol, rd_sim, sup, fut)]
d_tbl_fut <- dcast(
  d_tbl[, .(pr_decision = mean(fut)), keyby = .(N, rd_sim)],
  N ~ rd_sim, value.var = "pr_decision"
)

d_tbl <- rbind(
  cbind(descision = "Superiority", d_tbl_sup),
  cbind(descision = "Futility", d_tbl_fut)
)

gt(d_tbl) |>
  tab_options(
    table.font.size = "80%"
  ) 
```


```{r}
#| label: tidy-up
#| code-summary: Shut down timer
#| code-fold: true

toc(log = TRUE, quiet = TRUE)
```

Simulations run: `r unlist(tic.log(format = TRUE))`.

<!-- ::: {#refs} -->
<!-- ::: -->





